tstrait: a quantitative trait simulator for ancestral recombination graphs

### 2023-11-14 paper.tex
Updates to perf section

\section{Results}
\subsection{Phenotypic Model}
The additive model is a common assumption in GWAS
\citep{uffelmann2021}, and we assume that a quantitative trait $y$
is given by
\[
y=X\beta+\epsilon,
\]
where $\beta$ is the effect size, $X$ is a matrix describing the number
of causal mutations in the individual, and $\epsilon$ is the environmental
noise.
[TODO more discussion here of what this means in an ARG. Be clear about
causal mutations]
% Environmental noise is normally distributed with mean zero
% and a variance of $V_G(1-h^2)/{h^2}$,
% % distribution,
% % \begin{align}\label{eq:env}
%     % \epsilon\sim \mathcal{N}\left(0,\frac{V_G(1-h^2)}{h^2}\right),
% % \end{align}
% where $V_G$ is the additive genetic variance from the simulated effect sizes,
% and $h^2$ is the narrow-sense heritability provided as input by the user.
% The \texttt{tstrait} API is modular, however, and other forms of environmental
% noise can be easily generated by the user, if they wish.

@@

\subsection{Performance}
\texttt{Tstrait} is very efficient, and can be applied to datasets at the
largest scales.
Fig~\ref{fig:time}A shows how the trait simulation time scales
with number of individuals, for sequences of length 50, 100 and 200Mb.
For each sample size and sequence length,
we first simulate an ARG under a simple coalescent model with
human-like parameters using~\texttt{stdpopsim}~\citep{adrion2020}.
We then run a quantitative trait simulation
with 1000 causal sites on this ARG using \texttt{tstrait},
and report the mean CPU time over 10 independent replicates.
For comparison, we also report the CPU time required by a
modified version of the code used by \citet{martin2017} for a
subset of these simulated ARGs.
A comparison with existing methods that require reading
in genetic sequences in formats such as
VCF~\citep[e.g.][]{meyer2018,fernandes2020} is not feasible,
as the time required to read the data at this scale would be
far more than the minutes required here to perform
the simulations.
We do not report the memory usage here, but note that these
simulations for up to 5 million humans were run without difficult
on a laptop.

For each replicate of the simulation, \texttt{tstrait} is
used to simulate one quantitative trait with 1000 causal sites. (B)

@@

\begin{figure}[!t]%
\centering
% FIXME get the sizes right here and in the notebook where we're generating
% the figure so we don't have to scale here.
\includegraphics[width=213pt]{figures/time-scaling.pdf}
\caption{\textbf{Time taken to simulate quantitative traits.} Each point
represents the mean time for 10 independent runs under different random seeds.
The times reported are the total CPU time required to simulate quantitative
traits on an Intel(R) Core(TM) i9-11900H CPU and 16 GB of RAM. The trait model
is a normal distribution with $\mu=0$, $\sigma^2=1$, $h^2=0.3$, and
$\alpha=-1$. (A) Computational time with increasing number of individuals. The
whole genome dataset from \emph{Homo sapiens} with various number of
individuals and chromosome length are simulated by using \texttt{stdpopsim}
\citep{adrion2020}. For each replicate of the simulation, \texttt{tstrait} is
used to simulate one quantitative trait with 1000 causal sites. (B)
Computational time with increasing number of causal sites on two large
realistic ARGs. \texttt{tstrait} is used to simulate quantitative traits from
the simulated French Canadian dataset \citep{anderson2023} and the inferred
tree sequence dataset of the 1000 Genomes Project \citep{kelleher2019} with
varying number of causal sites. The French Canadian dataset is downloaded from
\url{https://zenodo.org/record/6839683}, and the inferred ARG from the 1000
Genomes Project is downloaded from \url{https://zenodo.org/record/3051855}.
Chromosome 9 is selected for both datasets.}\label{fig:time}
\end{figure}

@@

\section{Funding}

D.T. is supported by the Oxford Kobe Scholarship from the University of Oxford
and the Euretta J. Kellett Fellowship from Columbia University.

This work was supported by ...

@@

>>>>

\section{Results}
\subsection{Phenotypic Model}
The additive model is a common assumption in GWAS
\citep{uffelmann2021}, and we assume that a quantitative trait $y$
is given by
\[
y=X\beta+\epsilon,
\]
where $\beta$ is the effect size, $X$ is a matrix describing the number
of causal alleles in the individual, and $\epsilon$ is the environmental
noise.
[TODO more discussion here of what this means in an ARG. Be clear about
causal mutations]
% Environmental noise is normally distributed with mean zero
% and a variance of $V_G(1-h^2)/{h^2}$,
% % distribution,
% % \begin{align}\label{eq:env}
%     % \epsilon\sim \mathcal{N}\left(0,\frac{V_G(1-h^2)}{h^2}\right),
% % \end{align}
% where $V_G$ is the additive genetic variance from the simulated effect sizes,
% and $h^2$ is the narrow-sense heritability provided as input by the user.
% The \texttt{tstrait} API is modular, however, and other forms of environmental
% noise can be easily generated by the user, if they wish.

@@

\subsection{Performance}
\texttt{Tstrait} is very efficient, and can be applied to datasets at the
largest scales.
Supplementary Fig~\ref{fig:time}A shows how trait simulation time scales
with number of individuals, for sequences of length 50, 100 and 200Mb.
For each sample size and sequence length,
we first simulate an ARG under a simple coalescent model with
human-like parameters using~\texttt{stdpopsim}~\citep{adrion2020}.
We then run a quantitative trait simulation
with 1000 causal sites on this ARG using \texttt{tstrait},
and report the mean CPU time over 10 independent replicates.
For comparison, we also report the CPU time required by a
modified version of the code used by \citet{martin2017} for a
subset of these simulated ARGs.
A comparison with existing methods that require reading
in genetic sequences in formats such as
VCF~\citep[e.g.][]{meyer2018,fernandes2020} is not feasible,
as the VCFs involved are very large, and these tools assume that
the entire dataset fits in RAM.
[DISCUSS RAM USAGE]

@@

@@

\section{Funding}
DT is supported by the Oxford Kobe Scholarship from the University of Oxford
and the Euretta J. Kellett Fellowship from Columbia University.
JK acknowledges support from the Robertson Foundation,
NIH (research grants HG011395 and HG012473) and
EPSRC (research grant EP/X024881/1).

@@

\newpage
\renewcommand\thefigure{S\arabic{figure}}
\setcounter{figure}{0}
\renewcommand\thetable{S\arabic{table}}
\setcounter{table}{0}
\section{Supplementary Figures}

\begin{figure*}[t]%
\centering
% FIXME get the sizes right here and in the notebook where we're generating
% the figure so we don't have to scale here.
\includegraphics[width=213pt]{figures/time-scaling.pdf}
\caption{\textbf{Time taken to simulate quantitative traits.} Each point
represents the mean time for 10 independent runs under different random seeds.
The times reported are the total CPU time required to simulate quantitative
traits on an Intel(R) Core(TM) i9-11900H CPU and 16 GB of RAM. The trait model
is a normal distribution with $\mu=0$, $\sigma^2=1$, $h^2=0.3$, and
$\alpha=-1$. (A) CPU time with increasing number of individuals
based on ARGs simulated using \texttt{stdpopsim} with the [MODEL ID]
demographic mode. For each replicate we simulate
one quantitative trait with 1000 causal sites.
(B) CPU time with increasing number of causal sites on two large ARGs.
\texttt{tstrait} is used to simulate quantitative traits from
the simulated French Canadian dataset \citep{anderson2023} and the inferred
tree sequence dataset of the 1000 Genomes Project \citep{kelleher2019} with
varying number of causal sites. The French Canadian dataset is downloaded from
\url{https://zenodo.org/record/6839683}, and the inferred ARG from the 1000
Genomes Project is downloaded from \url{https://zenodo.org/record/3051855}.
Chromosome 9 is selected for both datasets.}\label{fig:time}
\end{figure*}
